The \dealii Library, Version 9.0

### 2018-04-25 9.0/paper.tex
Some edits.

\item Improved support for user parameters: a new \texttt{ParameterAcceptor}
  class has been added to the library. The class is intended to be used as a
  base for any class that wants to handle parameters using the
  \texttt{ParameterHandler} class. If you derive all your classes from
  \texttt{ParameterAcceptor}, and declare your parameters either with
  \texttt{parse/declare\_parameters} methods or via the
  \texttt{ParameterAcceptor::add\_parameter} method, then both the
  declaration and the
  parsing of parameter files are automatically managed by
  \texttt{ParameterAcceptor::initialize}, greatly simplifying
  dealing with parameters in user codes.

@@

Every function in the GridGenerator namespace now attaches a default manifold to
the curved parts of the domain, and sets reasonable defaults for manifold
indicators both in the domain and on the boundary, where appropriate. Manifold
classes are no longer sensible to settings made on boundary indicators, and are
only affected by manifold indicators, decoupling completely the concept of
boundary indicators and manifold indicator, leaving the former for boundary
conditions, and the latter for the geometrical description of the domain.

All functions that allowed to query for \texttt{Boundary} objects (deprecated
since version 8.5 of the \dealii library) have been removed, and have been
replaced by the equivalent methods that query \texttt{Manifold} classes.

The manifold smoothing algorithms applied in the Triangulation class and
MappingQGeneric have been changed from the old Laplace-style smoothing to a
transfinite interpolation that linearly blends between the descriptions on the
faces around a cell. The old transformation introduced boundary layers inside
cells that prevented convergence rates from exceeding \(3.5\) in the global
\(L^2\) errors on typical settings. This change also considerably improves mesh
quality on settings where curved descriptions are only applied to the boundary
rather than the whole volume.

A new manifold class \texttt{TransfiniteInterpolationManifold} implementing an
interpolation from a curved boundary description to the interior has been added.
This class enables high-order convergence rates of more than three in the power
of the mesh size for situations where a curved manifold can only be prescribed
to the boundary but not in a whole volume.

@@

\dealii has had a tutorial program (step-33) since 2007 that
demonstrates this technique based on the Trilinos Sacado \cite{Bartlett2006a} package, but
the functionality was not available pervasively throughout
deal.II. 
This has changed with release 9.0 where support is given for differentiation 
using a selection of ``white-listed'' libraries (namely ADOL-C \citep{Griewank1996a} 
and Sacado) and a subset of their supported number types. Currently, we offer support for
\begin{itemize}
\item ADOL-C taped (n-differentiable),
\item ADOL-C tapeless (once differentiable),
\item Sacado dynamic forward (once differentiable),
\item Sacado reverse (once differentiable),
\item Sacado nested dynamic forward (twice differentiable), and
\item Sacado nested reverse and dynamic forward (twice differentiable)
\end{itemize}
numbers. In practice, this support means that these ADOL-C and Sacado data
types can be used in the \texttt{FEValues}, \texttt{FEValuesViews},
\texttt{Tensor}, \texttt{SymmetricTensor},
and related classes that are generally used to assemble linear systems
and right hand sides.
Given the updated capabilities of the library, there is now a dedicated module 
that presents the AD compatibility and capabilities of the \dealii libraries. 
Furthermore, the use of Sacado is now  demonstrated in a much more simplified 
and transparent manner in an modernized version of an existing ``code gallery'' 
example \cite{Pelteret2016a}.

@@

The matrix-free infrastructure in \dealii{} was significantly overhauled for
the current release. The major new contribution is the support of face
integrals through a new class \texttt{FEFaceEvaluation}. The new class has a
similar interface as the previous \texttt{FEEvaluation}, and applies SIMD
vectorization over several faces in analogy to the intra-cell vectorization in
FEEvaluation. Discontinuous Galerkin operators are implemented defining two
face functions, one for interior and one for boundary faces, in addition to
the cell function. These kernels for the matrix-free operator evaluation are
now collected in the new function \texttt{MatrixFree::loop}. The data
structures have been particularly tuned for typical discontinuous Galerkin
setups involving operators with first and second spatial derivatives. Both
data access and computations have been thoroughly optimized and compared to
the performance boundaries of the hardware.

To give an example of the algorithmic improvements, the computation of the
values and gradients on all quadrature points for cell integrals has been
significantly improved, giving around 10--20\% better performance for the case
the kernels are compute bound. For the example of the reference cell gradient
of a solution field $\mathbf u$ in three space dimensions, the new release
applies the following change:
\begin{equation*}
  \text{previous: }
  \begin{bmatrix}
    D_{1} \otimes S_{2} \otimes S_{3}\\
    S_{1} \otimes D_{2} \otimes S_{3}\\
    S_{1} \otimes S_{2} \otimes D_{3}
  \end{bmatrix}
  \mathbf u
  \quad
  \leadsto
  \quad
  \text{new: }
  \begin{bmatrix}
    D_{1}^{\mathrm{co}} \otimes I_{2} \otimes I_{3}\\
    I_{1} \otimes D_{2}^{\mathrm{co}} \otimes I_{3}\\
    I_{1} \otimes I_{2} \otimes D_{3}^{\mathrm{co}}
  \end{bmatrix}
  \begin{bmatrix}
    S_{1} \otimes S_{2} \otimes S_{3}
  \end{bmatrix}
  \mathbf u.
\end{equation*}
The matrices $S_i$ contain the values of the one-dimensional shape functions
in one-dimensional quadrature points and $D_i$ their derivatives. When applied
with the usual sum factorization implementation described e.g.~in
\cite{KronbichlerKormann2012}, the old kernels amounted to 9 partial
summations---or rather 8 in the previous implementation of \dealii{} because
the application of $S_1$ for the $y$ and $z$ components of the gradient can be
merged. The new code performs a basis transformation to a related basis with
derivative matrix $D_i = D_i^{\mathrm{co}} S_i$, which is the basis of
Lagrange polynomials in the points of the quadrature. This change reduces the
number of partial sums to only 6 for the gradient, as the action of the unit
matrices $I_i$ needs not be implemented. In isolation, this spectral
element-like evaluation was previously available in \dealii{} for collocation
between nodal points and quadrature, but not used for general bases. A more
detailed description of the matrix-free modules is given in the preprint
\cite{KronbichlerKormann2017}.

@@

This release does not contain any new tutorial programs, though
several have been updated extensively for the changes to the manifold
handling as well as to adjust for current functionality
compared to what that was available when the programs were first written.

>>>>

\item Improved support for user-defined run-time parameters: a new \texttt{ParameterAcceptor}
  class has been added to the library. The class is intended to be used as a
  base for any class that wants to handle parameters using the
  \texttt{ParameterHandler} class. If you derive all your classes from
  \texttt{ParameterAcceptor}, and declare your parameters either with
  \texttt{parse/declare\_parameters} methods or via the
  \texttt{ParameterAcceptor::add\_parameter} method, then both the
  declaration and the
  parsing of parameter files are automatically managed by
  \texttt{ParameterAcceptor::initialize}, greatly simplifying
  dealing with parameters in user codes.

@@

\dealii{} has had the ability to attach \textit{manifold descriptions}
to all parts of a geometry for several releases already. These
descriptions are used when considering where to place new vertices
upon mesh refinement, in determining the mapping from reference cell
to real cell, and in a number of other contexts. However, for historical
reasons, manifold descripts have used some of the same code paths also
used for boundary indicators typically used to identify which parts of
the boundary correspond to what boundary conditions.

This connection has been severed in the current release: Boundary
indicators and manifold descriptions are now entirely separated. Furthermore,
all functions that allowed to query for old-style \texttt{Boundary} objects (deprecated
since version 8.5 of the \dealii library) have been removed, and have been
replaced by the equivalent methods that query \texttt{Manifold} classes.

There are also numerous improvements to the available manifold descriptions.
First, the manifold smoothing algorithms applied in the \texttt{Triangulation} class and
\texttt{MappingQGeneric} have been changed from the old Laplace-style smoothing to a
transfinite interpolation that linearly blends between the descriptions on the
faces around a cell. The old transformation introduced boundary layers inside
cells that prevented convergence rates from exceeding \(3.5\) in the global
\(L^2\) errors for typical settings. This change also considerably improves mesh
quality in situations where curved descriptions are only applied to the boundary
rather than the whole volume.

\marginpar{This paragraph seems to duplicate the previous. Can we
  merge the information?}
Secondly, a new manifold class
\texttt{TransfiniteInterpolationManifold} has been added that implements an
interpolation from a curved boundary description to the interior.
This class enables high-order convergence rates of better than ${\cal O}(h^3)$
for situations where a curved manifold can only be prescribed
to the boundary but not in a whole volume.

Finally,
every function in the \texttt{GridGenerator} namespace now attaches a default manifold to
the curved parts of the domain described by the generated mesh, and sets reasonable defaults for manifold
indicators both in the domain and on the boundary.

@@

\dealii has had a tutorial program (step-33) since 2007 that
demonstrates this technique based on the Trilinos Sacado \cite{Bartlett2006a} package, but
the functionality was not available pervasively throughout
deal.II. 
This has changed with release 9.0 where support is given for differentiation 
using a selection of ``white-listed'' libraries (namely ADOL-C \citep{Griewank1996a} 
and Sacado) and a subset of their supported number types. Currently,
we offer support for the following cases:
\begin{itemize}
\item ADOL-C taped (n-differentiable),
\item ADOL-C tapeless (once differentiable),
\item Sacado dynamic forward (once differentiable),
\item Sacado reverse (once differentiable),
\item Sacado nested dynamic forward (twice differentiable), and
\item Sacado nested reverse and dynamic forward (twice differentiable).
\end{itemize}
In practice, this support means that these ADOL-C and Sacado data
types can be used in the \texttt{FEValues}, \texttt{FEValuesViews},
\texttt{Tensor}, \texttt{SymmetricTensor},
and related classes that are generally used to assemble linear systems
and right hand sides.
Given the updated capabilities of the library, there is now a dedicated module 
that presents the AD compatibility and capabilities of the \dealii libraries. 
Furthermore, the use of Sacado is now  demonstrated in a much more simplified 
and transparent manner in an modernized version of an existing ``code gallery'' 
example \cite{Pelteret2016a}.

@@

The matrix-free infrastructure in \dealii{} was significantly overhauled for
the current release. The major new contribution is the support of face
integrals through a new class \texttt{FEFaceEvaluation}. The new class has a
similar interface as the existing \texttt{FEEvaluation} class, and applies SIMD
vectorization over several faces in analogy to the intra-cell vectorization in
FEEvaluation. Discontinuous Galerkin operators are implemented defining two
face functions, one for interior and one for boundary faces, in addition to
the cell function. These kernels for the matrix-free operator evaluation are
now collected in the new function \texttt{MatrixFree::loop}. The data
structures have been particularly tuned for typical discontinuous Galerkin
setups involving operators with first and second spatial derivatives. Both
data access and computations have been thoroughly optimized and compared to
the performance boundaries of the hardware.

To give an example of the algorithmic improvements, the computation of the
values and gradients on all quadrature points for cell integrals has been
significantly improved, yielding 10--20\% better performance for cases where
the kernels are compute bound. For the example of the reference cell gradient
of a solution field $\mathbf u$ in three space dimensions, the new release
applies the following change:
\begin{equation*}
  \text{previous: }
  \begin{bmatrix}
    D_{1} \otimes S_{2} \otimes S_{3}\\
    S_{1} \otimes D_{2} \otimes S_{3}\\
    S_{1} \otimes S_{2} \otimes D_{3}
  \end{bmatrix}
  \mathbf u
  \quad
  \leadsto
  \quad
  \text{new: }
  \begin{bmatrix}
    D_{1}^{\mathrm{co}} \otimes I_{2} \otimes I_{3}\\
    I_{1} \otimes D_{2}^{\mathrm{co}} \otimes I_{3}\\
    I_{1} \otimes I_{2} \otimes D_{3}^{\mathrm{co}}
  \end{bmatrix}
  \begin{bmatrix}
    S_{1} \otimes S_{2} \otimes S_{3}
  \end{bmatrix}
  \mathbf u.
\end{equation*}
The matrices $S_i$ contain the values of the one-dimensional shape functions
in one-dimensional quadrature points and $D_i$ their derivatives. When applied
with the usual sum factorization implementation described, for
example, in
\cite{KronbichlerKormann2012}, the old kernels amounted to 9 partial
summations -- or rather 8 in the previous implementation of \dealii{} because
the application of $S_1$ for the $y$ and $z$ components of the gradient can be
merged. The new code performs a basis transformation to a related basis with
derivative matrix $D_i = D_i^{\mathrm{co}} S_i$, which is the basis of
Lagrange polynomials in the points of the quadrature. This change reduces the
number of partial sums to only 6 for the gradient, as the action of the unit
matrices $I_i$ needs not be implemented. In isolation, this spectral
element-like evaluation was previously available in \dealii{} for collocation
between nodal points and quadrature, but not used for general bases. A more
detailed description of the matrix-free modules is given in the preprint
\cite{KronbichlerKormann2017}.

@@

This release does not contain any new tutorial programs, though
several have been updated extensively for the changes to the manifold
handling as well as to adjust for current functionality and coding styles
compared to what that was available when the programs were first written.
